---
title: "Data Visualization 01"
author: "Dr Saufi"
date: last-modified
format: 
  html:
    toc: true
    toc-title: Contents
    toc-location: left
    toc-depth: 5
    toc-expand: 1
    number-sections: true
    code-fold: true
    code-summary: "Code"
    code-link: true
    theme:
      light: united
      dark: cyborg
    css: styles.css
    self-contained: true
editor: visual
include-after-body: "footer.html"
---

# Prepare R Environment

Begin data analysis by creating a new project in RStudio to keep work organized. Steps to create a new project: Click File -\> Click New Project -\> Set name for the directory -\> Click Create Project

# Load Libraries

Load the necessary libraries for data visualization:

```{r}
# Load tidyverse for data manipulation and visualization
library(tidyverse)

# Load gapminder dataset for practice
library(gapminder)
```

# Basic Plots

Create different types of basic plots using ggplot2.

## Histogram

Visualize the distribution of life expectancy:

```{r}
ggplot(gapminder) +
  geom_histogram(aes(lifeExp))
```

## Barchart

Visualize the count of observations for each continent:

```{r}
ggplot(gapminder) +
  geom_bar(aes(continent))
```

## Line Graph

Visualize the change in life expectancy over time for each country:

```{r}
ggplot(gapminder) +
  geom_line(aes(x = year, y = lifeExp, group = country))
```

## Scatter Plot

Visualize the relationship between GDP per capita and life expectancy:

```{r}
ggplot(gapminder) +
  geom_point(aes(x = gdpPercap, y = lifeExp))
```

## Smooth Plot

Add a smooth trend line to the scatter plot to better visualize the relationship:

```{r}
ggplot(gapminder) +
  geom_smooth(aes(x = gdpPercap, y = lifeExp))
```

# Split Your Plots

Create faceted plots to split data visualizations by continent:

```{r}
ggplot(gapminder) +
  geom_histogram(aes(lifeExp)) +
  facet_wrap(~ continent)
```

# Saving Your Plots

Save your plots in different formats (PDF, PNG, JPG):

```{r}
# Save the plot as PDF
ggsave('myplot.pdf')

# Save the plot as PNG
ggsave('myplot.png')

# Save the plot as JPG
ggsave('myplot.jpg')
```

# References

Useful resources for further learning and customization:

-   [Different types of geom](https://ggplot2.tidyverse.org/reference/index.html)
-   [Customizing your plots](http://www.cookbook-r.com/Graphs/)
-   [More advanced R graphs](https://www.r-graph-gallery.com/)
