---
title: "Data Visualization 02"
author: "Dr Saufi"
date: last-modified
format: 
  html:
    toc: true
    toc-title: Contents
    toc-location: left
    toc-depth: 5
    toc-expand: 1
    number-sections: true
    code-fold: true
    code-summary: "Code"
    code-link: true
    theme:
      light: united
      dark: cyborg
    css: styles.css
    self-contained: true
editor: visual
include-after-body: "footer.html"
---

# Overview of Graphics Packages in R

There are several graphics packages in R designed for different purposes:

## Popular General Graphics Packages

1.  `graphics`: A base R package for basic plotting.
2.  `ggplot2`: A user-contributed package by Hadley Wickham, based on the grammar of graphics.
3.  `lattice`: Another user-contributed package for advanced plotting.

## Specialized Graphics Packages

1.  `survminer::ggsurvplot`: For survival analysis plots.
2.  `sjPlot`: For data visualization tailored to social sciences.

## Introduction to ggplot2

`ggplot2` is highlighted for its elegance, ease of use, and versatility: - Implements the grammar of graphics concept, making the learning process faster and facilitating the creation of complex graphics. - It combines the best features of base and lattice graphics without their drawbacks.

## Key Components of ggplot2

-   Start with: `ggplot()`
-   Specify data: `data = X`
-   Define variables: `aes(x = , y = )`
-   Choose graph type: `geom_histogram()`, `geom_point()`, etc.

## Questions to Ask Before Making Graphs

Before creating a graph, consider:

-   Which variable(s) to plot?
-   The type of variable(s) (factor or numerical)
-   The number of variables to plot together (single, two, or three variables)

## Preparation for Data Analysis

### Set a New Project Directory

Starting a new project in RStudio is recommended for clean and organized analysis:

1.  Steps to create a new project
2.  Go to File -\> Click New Project -\> New Directory -\> New Project -\> Create New Project

### Read Data

Common data formats include CSV, Excel, SPSS, Stata, and SAS files. Here’s an example of reading a CSV file:

```{r}
# Read a CSV file into R
mydata <- read.csv('HousingData.csv')
summary(mydata)
```

Packages for reading different data formats include `haven`:

-   SAS: `read_sas()` and `read_xpt()`
-   SPSS: `read_sav()` and `read_por()`
-   Stata: `read_dta()`

### Database Connections

Data from databases like MySQL, SQLite, Postgresql, and MariaDB are becoming increasingly important.

### Summary of ggplot2 Usage with Examples

Here's a basic guide and examples of using `ggplot2`:

#### Histogram

```{r}
library(ggplot2)
ggplot(mydata) +
  geom_histogram(aes(x = CRIM))

# x = variable_name
```

#### Scatter Plot

```{r}
ggplot(mydata) +
  geom_point(aes(x = CRIM, y = AGE))

# x = variable1, y = variable2
```

#### Line Graph

```{r}
ggplot(mydata) +
  geom_line(aes(x = CRIM, y = AGE, group = RAD))

# x = time_variable, y = measurement_variable, group = category_variable
```

#### Faceting

```{r}
ggplot(mydata) +
  geom_histogram(aes(x = CRIM)) +
  facet_wrap(~ RAD)

# x = variable_name, ~ = category_variable
```

# Hands-on 1: Packages

## Load the Packages

To create plots using ggplot2, we need to load the `tidyverse` package, which includes ggplot2 and other useful packages. Loading `tidyverse` provides access to various help pages, functions, and datasets included in the package suite.

```{r}
library(tidyverse)
```

## Open the Dataset

Use the `gapminder` dataset from the `gapminder` package. The dataset offers historical data on global life expectancy, GDP per capita, and population.

```{r}
# Load the gapminder package
library(gapminder)

# View the first few rows of the dataset
head(gapminder)
```

The `gapminder` dataset includes the following columns:

-   `country`: Country name (factor)
-   `continent`: Continent name (factor)
-   `year`: Year of observation (integer)
-   `lifeExp`: Life expectancy (numeric)
-   `pop`: Population (integer)
-   `gdpPercap`: GDP per capita (numeric)

```{r}
# Get a glimpse of the dataset structure
glimpse(gapminder)
```

The output provides a summary of the number of observations, types of variables, and a preview of the data:

-   1704 observations
-   6 variables (2 factor, 2 integer, and 2 numeric variables)

```{r}
# Summarize the dataset
summary(gapminder)
```

The summary function provides statistical insights into each variable:

-   Frequencies for factor variables (`country`, `continent`)
-   Minimum, 1st quartile, median, mean, 3rd quartile, and maximum values for numerical variables (`year`, `lifeExp`, `pop`, `gdpPercap`)

# Hands-on 2: Scatterplots

## Basic Scatterplot

Start by creating a basic scatter plot to visualize the relationship between the year and life expectancy in the gapminder dataset.

```{r}
# Load necessary libraries
library(tidyverse)
library(gapminder)

# Create a basic scatterplot
ggplot(data = gapminder) +
  geom_point(mapping = aes(x = year, y = lifeExp))
```

**Explanation:**

-   The `ggplot()` function initializes the plot and specifies the dataset (`gapminder`).
-   The `geom_point()` function creates a scatter plot with points representing the data.

**Observation:**

-   The plot shows an increase in life expectancy over the years, indicating a positive trend.

## Adding Another Variable

Enhance the scatter plot by adding a third variable to differentiate data points by continent.

```{r}
# Create a scatterplot with continent colors
ggplot(data = gapminder) +
  geom_point(mapping = aes(x = year, y = lifeExp, colour = continent))
```

**Explanation:**

-   The `aes()` function inside `geom_point()` now includes `colour = continent`, which adds color differentiation for each continent.

**Observation:**

-   European countries generally have higher life expectancy.
-   African countries tend to have lower life expectancy.
-   Outliers with very low life expectancy are noticeable in Asia and Africa.

#### Adding GDP as a Size Variable

Modify the scatter plot to reflect GDP per capita (`gdpPercap`) as the size of the points.

```{r}
# Create a scatterplot with GDP per capita as the size of points
ggplot(data = gapminder) +
  geom_point(mapping = aes(x = year, y = lifeExp, size = gdpPercap))
```

**Explanation:**

-   The `size = gdpPercap` aesthetic scales the size of the points according to GDP per capita.

**Observation:**

-   Countries with higher GDP generally have longer life expectancy.

#### Using Shape Instead of Color

You can use different shapes to represent continents when color is not an option, such as in black-and-white printouts.

```{r}
# Create a scatterplot with different shapes for continents
ggplot(data = gapminder) +
  geom_point(mapping = aes(x = year, y = lifeExp, shape = continent))
```

**Explanation:**

-   The `shape = continent` aesthetic assigns different shapes to data points based on continent.

#### Setting Color and Shape Outside `aes()`

You can also set a uniform color or shape for all points by placing the arguments outside the `aes()` function.

```{r}
# Create a scatterplot with all points colored blue
ggplot(data = gapminder) +
  geom_point(mapping = aes(x = year, y = lifeExp), colour = 'blue')
```

```{r}
# Create a scatterplot with all points shaped as plus signs
ggplot(data = gapminder) +
  geom_point(mapping = aes(x = year, y = lifeExp), shape = 3)
```

**Explanation:**

-   Setting `colour = 'blue'` or `shape = 3` outside `aes()` applies these attributes uniformly to all points.
-   The shape argument uses numbers to represent different symbols (`?pch` provides a list of shape codes).
-   Typing `?pch` in the R console will show you a reference for all the available point shapes and their corresponding codes.

# Hands-on 3: Subplots

To create subplots, we can use the `facet_wrap()` function to split our plots based on a factor variable. This allows for easy comparison of data subsets across different categories.

## Creating Subplots Based on Continents

### Create Subplots with 3 Rows

```{r}
ggplot(data = gapminder) +
  geom_point(mapping = aes(x = year, y = lifeExp)) + 
  facet_wrap(~ continent, nrow = 3)
```

**Explanation:**

-   `facet_wrap(~ continent)` splits the plot into subplots for each continent.
-   `nrow = 3` arranges the subplots into 3 rows.

### Create Subplots with 2 Rows

```{r}
ggplot(data = gapminder) +
  geom_point(mapping = aes(x = year, y = lifeExp)) + 
  facet_wrap(~ continent, nrow = 2)
```

**Explanation:**

-   Changing `nrow` to 2 arranges the subplots into 2 rows instead of 3, adjusting the layout for better readability based on your preferences or the amount of data.

## Customizing Subplots

-   You can further customize subplots by adjusting the number of columns (`ncol`), adding titles, modifying axis labels, and more to suit your analysis needs.
-   **Subplots facilitate comparison:** By breaking down the data into subplots based on the `continent` variable, it's easier to compare trends and patterns across different continents.
-   **Life expectancy trends:** Each subplot shows how life expectancy changes over time within each continent, making it clear how trends differ geographically.
-   By utilizing `facet_wrap()`, you can create organized and insightful visualizations that highlight differences and similarities across subsets of your data.

# Hands-on 4: Scatterplot, Smooth Plot, and Combining Plots

## Scatterplot and Smooth Plot

In `ggplot2`, different geometric objects (`geom_X()`) represent different visualizations.

### Scatterplot

```{r}
# Load necessary libraries
library(tidyverse)
library(gapminder)

# Create a scatterplot
ggplot(data = gapminder) +
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp))
```

**Explanation:**

-   `geom_point()` creates a scatter plot with GDP per capita on the x-axis and life expectancy on the y-axis.

### Smooth Plot

```{r}
ggplot(data = gapminder) +
  geom_smooth(mapping = aes(x = gdpPercap, y = lifeExp))
```

**Explanation:**

-   `geom_smooth()` adds a smooth line to the plot, indicating trends in the data.

We can generate the smooth plot based on continent using the `linetype()`. We use `log(gdpPercap)` to reduce the skewness of the data.

```{r}
ggplot(data = gapminder) +
  geom_smooth(mapping = aes(x = log(gdpPercap), y = lifeExp, linetype = continent))
```

**Explanation:**

-   `aes(x = log(gdpPercap), y = lifeExp, linetype = continent)` uses the logarithm of GDP per capita to reduce skewness and differentiates lines by continent.

We can also generate the smooth plot using colour.

```{r}
ggplot(data = gapminder) +
  geom_smooth(mapping = aes(x = log(gdpPercap), y = lifeExp, colour = continent))
```

**Explanation:**

-   `colour = continent` differentiates lines by continent using color.

## Combining Geoms

We can overlay multiple geoms (geometric objects) in a single plot to visualize multiple aspects of the data simultaneously.

```{r}
ggplot(data = gapminder) +
  geom_point(mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_smooth(mapping = aes(x = log(gdpPercap), y = lifeExp))
```

**Explanation:**

-   This combines a scatterplot and a smooth plot, but it repeats the `mapping` for each geom.

The codes above show duplication or repetition. To avoid this, we can pass the mapping to `ggplot()`.

```{r}
ggplot(data = gapminder, mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point() +
  geom_smooth()
```

**Explanation:**

-   Passing the `mapping` to `ggplot()` applies it to all geoms, reducing code duplication.

And we can expand this to make scatterplot with different colors for continents.

```{r}
ggplot(data = gapminder, mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point(mapping = aes(colour = continent)) +
  geom_smooth()
```

**Explanation:**

-   This plot combines a scatterplot with points colored by continent and a smooth line.

Or expand this to make the smooth plot shows different colour for continent.

```{r}
ggplot(data = gapminder, mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point() +
  geom_smooth(mapping = aes(colour = continent))
```

**Explanation:**

-   This plot combines a scatterplot and a smooth plot, with the smooth lines colored by continent.

Or combining both scatterplot and smooth plot.

```{r}
ggplot(data = gapminder, mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point(mapping = aes(shape = continent)) +
  geom_smooth(mapping = aes(colour = continent))
```

**Explanation:**

-   This plot uses different shapes for continents in the scatterplot and different colors for continents in the smooth plot.

## Summary

-   **Scatterplot (`geom_point()`):** Visualizes individual data points.
-   **Smooth Plot (`geom_smooth()`):** Adds trend lines to visualize the overall trend.
-   **Combining Geoms:** Overlay different types of plots to provide more comprehensive visual analysis.
-   **Aesthetics Mapping:** Use `aes()` to map data variables to visual properties like color, shape, and size.
-   **Avoiding Duplication:** Pass common aesthetics mapping to `ggplot()` to avoid repetition and simplify code.

# Hands-on 5: Bar Plot and Histogram

## Bar Plot

A bar plot is useful for displaying the distribution of categorical data. Here, bar plots are created to show the frequency and proportion of continents in the `gapminder` dataset.

**Bar Plot Showing Frequency**

```{r}
# Load necessary libraries
library(tidyverse)
library(gapminder)

# Create a bar plot showing frequency of each continent
ggplot(data = gapminder) +
  geom_bar(mapping = aes(x = continent))
```

**Explanation:**

-   `geom_bar()` creates a bar plot with continents on the x-axis and the count (frequency) of records on the y-axis.

**Bar Plot Showing Proportion**

```{r}
# Create a bar plot showing proportion of each continent
ggplot(data = gapminder) +
  geom_bar(mapping = aes(x = continent, y = ..prop.., group = 1))
```

**Explanation:**

-   `y = ..prop..` scales the y-axis to show the proportion of records for each continent instead of the count.
-   `group = 1` ensures that proportions are calculated for the entire dataset.

## Histogram

Histograms are used for visualizing the distribution of numerical data. Here, we create a histogram for life expectancy in the `gapminder` dataset.

```{r}
# Create a histogram for life expectancy
ggplot(data = gapminder, aes(x = lifeExp)) + 
  geom_histogram()
## `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
```

**Explanation:**

-   `geom_histogram()` creates a histogram with life expectancy on the x-axis.
-   By default, `geom_histogram()` uses 30 bins. You can adjust the `binwidth` parameter to get a more detailed or summarized view of the data distribution.

## Summary

-   **Bar Plot (`geom_bar()`):** Visualizes the distribution of categorical data. It can show frequencies or proportions.
-   **Histogram (`geom_histogram()`):** Visualizes the distribution of numerical data by dividing the data into bins and counting the number of observations in each bin.
-   **Proportional Bar Plot:** Use `..prop..` in the y aesthetic to display proportions instead of counts.

# Hands-on 6: Meaningful and Beautiful Plots

## Customizing Title

Customizing plot titles can make your visualizations more informative.

**Creating the Initial Plot**

```{r}
# Load necessary libraries
library(tidyverse)
library(gapminder)

# Create a scatter plot with a smooth line
mypop <- ggplot(data = gapminder, mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point() +
  geom_smooth(mapping = aes(colour = continent))
mypop
```

**Adding a Title**

```{r}
mypop + 
  ggtitle("Scatterplot showing the relationship of GDP in log and life expectancy")
```

**Adding a Title with Multiple Lines**

```{r}
mypop + 
  ggtitle("Scatterplot showing the relationship of GDP in log and life expectancy:\nData from Gapminder")
```

## Adjusting Axes

Adjusting axis tick marks and labels can enhance readability.

**Specifying Tick Marks**

-   min = 0
-   max = 12
-   interval = 1

```{r}
mypop + 
  scale_x_continuous(breaks = seq(0, 12, 1))
```

**Labeling Axes** Add axis labels and a title with multiple lines.

```{r}
mypop + 
  ggtitle("Scatterplot showing the relationship of GDP in log and life expectancy:\nData from Gapminder") + 
  ylab("Life Expectancy") + 
  xlab("Per Capita GDP (log scale)")
```

## Choosing Themes

`ggplot2` provides several themes to change the overall appearance of plots:

**Default Theme (Gray)**

-   The default theme in ggplot2 is `theme_gray()`.

**Black and White Theme**

```{r}
mypop + theme_bw()
```

**Classic Theme**

```{r}
mypop + theme_classic()
```

## Summary

-   **Customizing Titles:** Adding informative titles and subtitles helps in understanding the context of the plot.
-   **Adjusting Axes:** Customizing tick marks and axis labels makes the plot more readable.
-   **Choosing Themes:** Different themes can enhance the aesthetic appeal of the plot and highlight different aspects of the data.

# Hands-on 7: Saving Plots

## Preferred Format for Saving

The preferred format for saving plots in R is often PDF because it preserves the quality and scalability of the graphics. However, other formats like PNG and JPG are also commonly used.

## Saving Plots Using ggplot2

**Creating a Complete Plot**

```{r}
# Load necessary libraries
library(tidyverse)
library(gapminder)

# Create the plot
mypop <- ggplot(data = gapminder, mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point() +
  geom_smooth(mapping = aes(colour = continent)) +
  ggtitle("Scatterplot showing the relationship of GDP in log and life expectancy:\nData from Gapminder") +
  ylab("Life Expectancy") + 
  xlab("Per Capita GDP (log scale)") +
  scale_x_continuous(breaks = seq(0, 12, 1)) +
  theme_classic()

# Display the plot
mypop
```

**Saving the Plot**

-   You can save the current plot displayed on the screen to various file formats using `ggsave()`.

```{r}
# Save the plot as a PDF
ggsave("my_pdf_plot.pdf")
```

```{r}
# Save the plot as a PNG
ggsave("my_png_plot.png")
```

```{r}
# Save the plot as a JPG
ggsave("my_jpg_plot.jpg")
```

**Customizing the Saved Plot**

-   The dimensions and resolution of the saved plot can also be customized.

```{r}
# Save the plot as a customized PDF
ggsave("my_pdf_plot2.pdf", width = 10, height = 6, units = "cm")
```

```{r}
# Save the plot as a customized PNG
ggsave("my_png_plot2.png", width = 10, height = 6, units = "cm")
```

```{r}
# Save the plot as a customized JPG
ggsave("my_jpg_plot2.jpg", width = 10, height = 6, units = "cm")
```

## Summary

-   **Preferred Format:** PDF is often preferred for its quality and scalability, but PNG and JPG are also useful.
-   **Saving Plots:** Use `ggsave()` to save plots in different formats.
-   **Customization:** Customize the size and resolution of the saved plot by specifying the `width`, `height`, and `units` in the `ggsave()` function.

# References

-   Cookbook for R: A collection of R recipes for visualization and analysis. <http://www.cookbook-r.com/Graphs/>
-   R Graph Gallery: A collection of charts made with the R programming language. <https://www.r-graph-gallery.com/>
